# Exercise 3: Analysis


## TASK 2
## Reading the data into R using the link provided:

```{r}
alc_data <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt", header=TRUE, sep=",")
str(alc_data)
```
 *The dataset has been downloaded from UCI Machine Learning Depository. The data have been collected from secondary education students in two Portuguese schools by using school reports and questionnaires. The data contain 35 variables that cover the students' grades, democraphic, social and school-related measures.*
 
## TASK 3

*The aim of the current analyses is to  study the relationships between high/non-high alcohol
consumption (high_use) and some of the other variables in the data.*


## Relationship of alcohol consumption with sex, family relationships, going out with friends and health status

*I chose the following 4 variables of interest for the analysis:*
freetime - free time after school (numeric: from 1 - very low to 5 - very high)
famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent);
goout - going out with friends (numeric: from 1 - very low to 5 - very high);
health - current health status (numeric: from 1 - very bad to 5 - very good).

*My hypotheses are:*
1) high alcohol consumption is more probable in cases of having more freetime
2) high alcohol consumption is more probable in cases of bad family relationships vs. good relationships
3) high alcohol consumption is more probable in students who tend to go out with friends
4) high alcohol consumption is more probable in cases of worse current health status


## TASK 4

First checking how many a student uses non-high vs. high amounts of alcohol:
```{r}
library(dplyr); library(ggplot2); library(psych)
alc_data %>% group_by(high_use) %>% summarise(count = n())
```
In this dataset, n=270 students have non-high, n=112 have high alcohol consumption.

Group frequencies of the 4 chosen variables:

```{r}
table(alc_data$freetime)
table(alc_data$famrel)
table(alc_data$goout)
table(alc_data$health)
```


How the use of alcohol is divided among the groups of the 4 variables of interest:

```{r}
cross1 <- table(alc_data$high_use, alc_data$freetime)
round(prop.table(cross1,2)*100, digits=0) #percentages in each group

cross2 <- table(alc_data$high_use, alc_data$famrel)
round(prop.table(cross2,2)*100, digits=0)

cross3 <- table(alc_data$high_use, alc_data$goout)
round(prop.table(cross3,2)*100, digits=0)

cross4 <- table(alc_data$high_use, alc_data$health)
round(prop.table(cross4,2)*100, digits=0)
```

Based on these contingency tables, it seems that a) as freetime increases (1->5), the probability of high alcohol consumption increases; b) as the quality of family relationships increases (1->5), the probability of high alcohol consumption first increases, then decreases; c) as going out with friends increases (1->5), the probability of high alcohol consumption increases; and d) as current health status increases (1->5), the probability of high alcohol consumption doesn't change.

*My hypotheses were:*
1) high alcohol consumption is more probable in cases of having more freetime
2) high alcohol consumption is more probable in cases of bad family relationships vs. good relationships
3) high alcohol consumption is more probable in students who tend to go out with friends
4) high alcohol consumption is more probable in cases of worse current health status

This preliminary exploration seems to support my hypotheses 1) and 3), but not 2) and 4).

It is worth noting that the percentages can be misleading with very small number of cases (e.g. famrel=1, n=9).

##TASK 5

##Logistic regression models

Using glm() to fit a logistic regression model; computing odds ratios (OR) and 95% confidence intervals (CI) of the four variables of interest:

```{r}
m_1 <- glm(high_use ~ freetime + famrel + goout + health, data = alc_data)
OR_1 <- coef(m_1) %>% exp 
CI_1 <- confint(m_1) %>% exp

summary(m_1)
cbind(OR_1, CI_1)

```

Odds ratio refers to the ratio of two odds (in this case, the odds of having high/non-high alcohol consumption). OR = 1 means these two conditions are just as likely. The confidence interval shows the range the OR falls into with 95% probability (default).

According to this model, having a) more or less freetime or b) better or worse current health status makes no difference in terms of alcohol consumption (CI includes one). 

On the other hand, both the quality of family relationships and the frequency of going out with friends seem to be valid predictors of the probability of high alcohol consumption (p<0.001 for both). Here, the first level of the outcome variable (FALSE) denotes failure and the other (TRUE) success. Because higher odds = higher probability of "success" and lower odds = lower probability of "success" vs. "failure", it seems that better quality of family relationships is associated with slightly lower probability of high alcohol consumption. Conversely, more frequent going out with friends is associated with slightly higher probability of high alcohol consumption. These results support my hypotheses 2) and 3), but not 1) and 4).


##The model with only statistically significant predictors included:
```{r}
m_2 <- glm(high_use ~ famrel + goout, data = alc_data, family = "binomial")

#Analyzing the predictive power of the model:
prob <- predict(m_2, type = "response") #Creating object prob that indicates the predicted probabilities on the scale of the response var (high_use)
alc_data <- mutate(alc_data, probability = prob)  #Adding a new variable called probability to the dataset
alc_data <- mutate(alc_data, prediction = probability > 0.5) #Adding a new variable called prediction; TRUE if probability > 0.5

str(alc_data) #Checking how the dataset looks

```

```{r}
#Predictions vs. actual values of alcohol consumption:
table(high_use = alc_data$high_use, prediction = alc_data$prediction)
```

Total proportions of inaccurately classified individuals:
```{r}
proportion_inacc <- (64 + 27) / 382
proportion_inacc
```


Based on the results, the model makes an inaccurate prediction in 23.8% of the cases.


```{r}
proportion_acc_false <- 243 / (243 + 64)
proportion_acc_false
```
```{r}
proportion_acc_true <- 48 / (48 + 27)
proportion_acc_true
```


The model predicts FALSE cases (low alcohol consumption) at 79.2% accuracy and TRUE cases (high alcohol consumption) at 64% accuracy. Therefore, the model is a more accurate predictor than, say, tossing a coin. However, this is surely not the best possible model to predict the probability of high alcohol use in this population.